---
title: Tidy Pandas
author: R package build
date: '2022-02-20'
slug: tidy-pandas
categories:
  - R
tags:
  - R
  - tidyverse
  - pandas
  - python
subtitle: ''
summary: ''
authors: []
lastmod: '2022-02-20T16:52:19+11:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output: 
  html_document: 
    toc: yes
    self_contained: no
---

Uni starts back up soon so I thought it would be a good idea to brush up on my `python`. I have avoided using `python` since `R` is much more user friendly between `dplyr` and `ggplot2`. Now that RStudio has `python` computability via `reticulate` there is not really a good reason to completely avoid `python`. I think it is always good practice to get more comfortable with different languages, because sometimes you will need to do a task that is only available using a specific tool. Similarly, there are times when a group project works best when everybody is able to use a similar language.

As tempting as it might be to do all my data wrangling via `tidyverse`, I have been practising using pandas. This post was entirely written in RStudio, however, the `python` code will run in a notebook alternative such as Jupyter or VS Code. 

### Installing Packages {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
install.packages("ggplot2")
install.packages("dplyr")
install.packages("purrr")

#or simply
install.packages("tidyverse")
```

#### Python

```{python eval=FALSE}
!pip3 install seaborn
!pip3 install numpy
!pip3 install pandas

#You can also install these packages in the terminal 
pip3 install seaborn
pip3 install numpy
pip3 install pandas
```

Learning `python` was overwhelming because you use the terminal much more often than you need to when using `R`. Over time, I have began to appreciate using a virtual environment because you can easily run multiple versions of `python`, which makes it more practical when you are using a package that requires an older version. 

### Importing Data, Loading Libraries {.tabset .tabset-fade .tabset-pills}

I have hidden the output of the code, however, you can view it by clicking the dropdown menu. I did this primarily for myself so it was easier to scroll down, but I think it is also more practical since the output is not necessarily the focus of this post.

#### R

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

df <- diamonds

df %>% head()
```

#### Python

```{python message=FALSE, warning=FALSE}
import seaborn as sns
import numpy as np
import pandas as pd
sns.set_style('white')

df = sns.load_dataset('diamonds')

df.head()
```

### Example of Functions {.tabset .tabset-fade .tabset-pills}

#### Verbs

|                            dplyr |                           pandas |
|---------------------------------:|---------------------------------:|
|         `filter()` and `slice()` |  `query()` and `loc[]`, `iloc[]` |
|                      `arrange()` | `sort_values` and `sort_index()` |
|        `select()` and `rename()` |     `__getitem__` and `rename()` |
|                       `select()` |                       `filter()` |
|                     `distinct()` |              `drop_duplicates()` |
|                       `mutate()` |                         `assign` |
|                    `summarise()` |                            `agg` |
|                     `group_by()` |                      `groupby()` |
| `sample_n()` and `sample_frac()` |                         `sample` |
|                            `%>%` |                       `pipe[^1]` |

#### Example

|                                           dplyr |                                       pandas |
|------------------------------------------------:|---------------------------------------------:|
|                 `filter(df, col == 'val')`      |                   `df.query('col == "val"')` |
|                         `arrange(df, col)`      |                      `df.sort_values('val')` |
|           `rename(df, new_name = old_name)`     | `df.rename(columns = {old_name = new_name})` |
|                           `select(df, col)`     |                              `df.loc['val']` |
|           `distinct(df, col, .keep_all = TRUE)` |              `df[['val']].drop_duplicates()` |
|                  `mutate(new_var = col - col2)` |      `df.assign(new_var = df.col - df.col2)` |
| `summarise(mean = mean(col2), n = count(col1))` |  `df.agg({"col1": "count", "col2", "mean"})` |
|                             `group_by(df, col)` |                          `df.groupby('col')` |
|                                           `%>%` |                                   `pipe[^1]` |

[^1]: To the best of my knowledge, you can still pipe without using the function, however, I have not explored it that much.

One of the confusing things are first is that there are many similar functions under different names. I personally find it easier to remember them by the way I write my code. For example, by only using `<-` as an assignment operator in `R`, I find it easier to treat the two languages differently. 

## `select()` 

### selecting columns {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
select(df, color, cut)

#or

df %>%
  select(color, cut)
```

<details>
<summary>View Output</summary>

```{r}
select(df, color, cut)
```

</details>

#### Python

```{python eval=FALSE}
df.filter(['color', 'cut'])

#or

df[['x', 'y', 'z']]
```

<details>
<summary>View Output</summary>

```{python}
df.filter(['color', 'cut'])
```

</details>

### If we want to select a range of columns {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
select(df, x:z)

#or

df %>% select(x:z)
```

<details>
<summary>View Output</summary>

```{r}
select(df, x:z)
```

</details>

#### Python

```{python eval=FALSE}
df.loc[:, 'x':'z']
```

<details>
<summary>View Output</summary>

```{python}
df.loc[:, 'x':'z']
```

</details>

### If we want to pipe it {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
df %>% 
  select(color, cut)
```

<details>
<summary>View Output</summary>

```{r}
df %>% 
  select(color, cut)
```

</details>

#### Python

```{python eval=FALSE}
(df
.filter(['color', 'cut'])
)
```

<details>
<summary>View Output</summary>

```{python}
(df
.filter(['color', 'cut'])
)
```

</details>

### If we want to drop a certain column {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
select(df, -(x:z))

#or

df %>%
  select(-(x:z))
```

<details>
<summary>View Output</summary>

```{r}
select(df, -(x:z))
```

</details>


#### Python

```{python eval=FALSE}
(df
.drop(['x', 'y', 'z'], axis = 1)
)
```

<details>
<summary>View Output</summary>

```{python}
(df
.drop(['x', 'y', 'z'], axis = 1)
)
```

</details>


## `filter()`

### filtering on one condition {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
filter(df, color == 'E')
```

<details>
<summary>View Output</summary>

```{r}
filter(df, color == 'E')
```

</details>


#### Python

```{python eval=FALSE}
(df
.query("color == 'E'")
)
```

<details>
<summary>View Output</summary>

```{python}
(df
.query("color == 'E'")
)
```

</details>

### If we want multiple conditions {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
filter(df, color == 'E', cut == 'Good')

#or

filter(df, color == 'E' & cut == 'Good')
```

<details>
<summary>View Output</summary>

```{r}
filter(df, color == 'E', cut == 'Good')
```

</details>

#### Python

```{python eval=FALSE}
(df
.query('color == "E" & cut == "Good"')
)
```

<details>
<summary>View Output</summary>

```{python}
(df
.query('color == "E" & cut == "Good"')
)
```

</details>

### If we want multiple conditions in one column {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
df %>% 
    filter(color %in% c('E', 'J'))
```

<details>
<summary>View Output</summary>

```{r}
df %>% 
    filter(color %in% c('E', 'J'))
```

</details>

#### Python

```{python eval=FALSE}
(df
.query('color in ["E", "J"]')
)
```

<details>
<summary>View Output</summary>

```{python}
(df
.query('color in ["E", "J"]')
)
```

</details>

### Count Missing Values {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
# sum of missing values in each column
df %>% 
  summarise(across(everything(), ~sum(is.na(.))))

purrr::map_df(df, ~sum(is.na(.)))
```

<details>
<summary>View Output</summary>

```{r}
purrr::map_df(df, ~sum(is.na(.)))
```

</details>

#### Python

```{python eval=FALSE}
df.isna().sum()
```

<details>
<summary>View Output</summary>

```{python}
df.isna().sum()
```

</details>

### Count Unique Values in Each Column {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
# getting the count of unique values in each column 
df %>% 
  summarise(across(everything(), n_distinct))

#can also map across for the same result
purrr::map_df(df, ~sum(n_distinct(.)))

# if you just want numerical columns
df %>% 
  summarise(across(where(is.numeric), n_distinct))
```

<details>
<summary>View Output</summary>

```{r}
purrr::map_df(df, ~sum(n_distinct(.)))
```

</details>

#### Python


```{python, eval = F}
df.nunique()

# If you want unique values in numeric columns
df.select_dtypes(include = np.number).nunique()
#or
df.select_dtypes('number').nunique()

# If you just want the column names of numeric type
df.select_dtypes('number').columns
# If you want them as a list
df.select_dtypes(include = np.number).columns.tolist()

# count and unique values
df.agg(['count', 'size', 'nunique'])

#for the proportions
df.select_dtypes(include = np.number).value_counts(normalize = True)
#or
df.select_dtypes('number').value_counts(normalize = True)
```

<details>
<summary>View Output</summary>

```{python}
(df
.select_dtypes(include = np.number)
.nunique()
)
```

</details>

### Complex Pipings {.tabset .tabset-fade .tabset-pills}

#### R

```{r eval=FALSE}
df %>%
  select(starts_with('c')) %>%
  filter(cut %in% c('Ideal', 'Premium')) %>%
  group_by(cut, color, clarity) %>%
  summarise(
    avgcarat = mean(carat, na.rm=TRUE),
    n = n()
    ) %>%
  arrange(-avgcarat) %>% #desc(avgcarat) also works
  head()
```

<details>
<summary>View Output</summary>

```{r}
df %>%
  select(starts_with('c')) %>%
  filter(cut %in% c('Ideal', 'Premium')) %>%
  group_by(cut, color, clarity) %>%
  summarise(
    avgcarat = mean(carat, na.rm=TRUE),
    n = n()
    ) %>%
  arrange(-avgcarat) %>% #desc(avgcarat) also works
  head()
```

</details>


#### Python

```{python eval=FALSE}
(df
 .filter(regex = '^c')
 .query('cut in ["Ideal", "Premium"]')
 .groupby(['cut', 'color', 'clarity'])
 .agg(['mean', 'size'])
 .sort_values(by = ('carat', 'mean'), ascending = False)
 .head())
```

<details>
<summary>View Output</summary>

```{python}
(df
 .filter(regex = '^c')
 .query('cut in ["Ideal", "Premium"]')
 .groupby(['cut', 'color', 'clarity'])
 .agg(['mean', 'size'])
 .sort_values(by = ('carat', 'mean'), ascending = False)
 .head())
```

</details>

### More Examples {.tabset .tabset-fade .tabset-pills}

#### Transforming

| R                                | pandas                                                |
|----------------------------------|-------------------------------------------------------|
| `select(df, col_one = col1)`     | `df.rename(columns = {'col1': 'col_one'})['col_one']` |
| `rename(df, col_one = col1)[^2]` | `df.rename(columns = {'col1': 'col_one'})`            |
| `mutate(df, c = a - b)`          | `df.assign(c = df['a'] - df['b'])`                    |

[^2]: You can achieve the same result with `select()`. However, `rename()` can be helpful if you do not want to drop or add, or relocate columns.

#### Sorting

| R                                | pandas                                      |
|----------------------------------|---------------------------------------------|
| `arrange(df, col1, col2)`        | `df.sort_values(['col1', 'col2'])`          |
| `arrange(df, desc(col1))[^3]`    | `df.sort_values('col1', ascending = False)` |

[^3]: I personally prefer using `arrange(df, -col1)`

#### Grouping and Summarising

| R                                                                  | pandas                                      |
|--------------------------------------------------------------------|---------------------------------------------|
| `summary(df)`                                                      | `df.describe()`                             |
| `group_by(df, col1)`                                               | `df.groupby('col1')`                        |
| `group_by(df, col1) %>% summarise(avg = mean(col1, na.rm = TRUE))` | `df.groupby('col1').agg({'col1' : 'mean'})` |
| `group_by(df, col1) %>% summarise(total = sum(col1))`              | `df.groupby('col1').sum()`                  |


